Memes: an R interface to the MEME Suite

### 2020-11-05 manuscript.Rmd
more edits

Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, for example,
identification of kinase substrates based on similarities in phosphorylation
site sequence, or characterizing DNA elements based on shared transcription
factor binding sequences (TODO: ref kinase motif:
https://doi.org/10.1016/0968-0004(90)90073-K). Thus the ability to identify,
classify, and compare motifs represents a key step in the analysis of biological
sequences. **TODO: R/Bioconductor transition sentence**. Although tools that
implement these features exist in Bioconductor, they often rely on
proprietary data types or have poor interoperability with other tools, creating
barriers to comprehensive motif analysis within R. The recent addition of
`universalmotif` to the Bioconductor landscape partially addresss these issues
by standardizing motif formats and providing a powerful set of tools for
examining the properties of motifs, however the Bioconductor motif analysis
landscape remains incomplete \cite{universalmotif}.

The MEME Suite is a series of commandline and web-based tools that provide a variety
of features for motif analysis such as *de-novo* motif discovery, enrichment
testing, comparison, and matching \cite{meme-suite}. The MEME Suite is a widely
utilized set of tools, with over 22,000 citations to date, and greater than 30,000 unique
users of the webserver implementation annually \cite{meme-usage-report-page,
google-scholar-meme}. The MEME Suite family of tools provides a comprehensive
set of utilities to interrogate motif content across a broad range of biological
contexts. However, several factors limit the full potential of these tools for
use in data analysis. MEME Suite tools require carefully formatted inputs to
achieve full functionality, yet few tools exist to simplify the process of data
formatting, requiring instead that users write custom code to prepare their
data, or prepare the inputs by hand, both of which have the potential to be
error prone without rigorous testing. Furthermore, the output data from each
MEME Suite tool often have complex structures that must be parsed to extract the
full suite of information, again requiring users to write custom code for this
task. Finally, although the data-generation capabilities of the MEME Suite are
excellent, the tools lack powerful ways to visualize the results. Collectively,
these factors act as barriers to adoption and preclude deep analysis of the
data, as well as communication of results to the scientific community.

@@

`memes` also provides simple utilities for data visualization where appropriate.
Providing R representations of MEME Suite data objects enables generation of
novel data visualizations using the well established tools available in the R
landscape. Therefore, `memes` empowers users to deeply analyze and visualize
motif analysis data.

@@

MEME tools return HTML reports for each tool that display data in a
user-friendly way, however, these files are not ideal for downstream processing.
Depending on the tool, data are also returned in tab-separated, or XML format,
which are more amenable to computational processing. However, in the case that
tab-separated data are returned, results are often incomplete. For example, the
TomTom tool, which compares query motifs to a database of known motifs, returns
tab-separated results that do not contain the frequency matrix representations
of the matched motifs. Instead, users must write custom code to parse these
matrices back out from the input databases, creating additional barriers to
analysis. In the case that data are returned in XML format, these files contain
all relevant data, however are difficult to parse, again requiring users to
write custom code to extract the necessary data. Finally, the data types
contained in MEME Suite outputs are complex and multidimensional, and thus
require special data structures to properly organize the data in memory.

@@

Data returned from MEME Suite tools are complex and multidimensional, making
them difficult to represent in a simple data structure. For example, MEME and
DREME return de-novo discovered motifs from query sequences along with
statistical information about their enrichment. In this instance, a position
frequency matrix (PFM) of the discovered motif can be represented as a matrix,
while the properties of that matrix (e.g. the name of the motif, the E-value
from the enrichment test, background nucleotide frequency, etc.) must be encoded
outside of the matrix, while maintaining their relationship to the
corresponding PFM. However, other MEME Suite tools, such as TOMTOM, which
compares one motif to a series of several motifs to identify possible matches,
produce an additional layer of complexity such that input matrices (which
contain metadata as previously described) can have a one-to-many relationship
with other motif matrices (again with their own metadata). Thus an ideal
representation for these data is one which can hold an unlimited number of motif
matrices and their metadata, contain complex hierarchies, and be easily
manipulated using standard analysis tools.

The `universalmotif` data structure is a powerful R/Bioconductor representation
for motif matrices and their associated metadata \cite{universalmotif}. However,
`universalmotif` objects are not ideal for performing bulk manipulation or
queries of metadata, nor do they support complex data structures such as nested
elements. R data.frames support nesting arbitrary data types within columns, but
are not sufficient to properly store motif matrices. In order to create an
extensible, nested data structure for tracking MEME Suite data, `memes`
leverages the power of R data.frames and the `universalmotif` format by nesting
`universalmotif` objects within data.frames. These universalmotif data.frames
store information about motifs along rows, where each column stores metadata
associated with a motif, including a `motif` column holding a `universalmotif`
representation of that motif, which can be manipulated by altering columns of
the data.frame (described in detail below). Adding one-to-many relationships within this
structure is trivial, as additional unviersalmotif data.frames can be nested
within eachother. These universalmotif data.frames form the basis for a majority
of `memes` data outputs. These structures are also valid input types to `memes`
functions, and when used as such, output data are appended as new columns to the
input data, ensuring data provenance. Finally, because the universalmotif
data.frame is just a data.frame, it can be manipulated using base R or
`tidyverse` functions, just like regular R data.frames, lowering the cognitive
overhead to interfacing with these structures.

@@

Discriminative (or "differential") motif analysis, in which motifs are discovered in one set of
sequences relative to another set, can uncover biologically relevant motifs
associated with membership to distinct categories. For example, during analysis
of multiomic data, users can identify different functional categories of genomic
regions through integration with orthogonal datasets (e.g.  categorizing
transcription factor binding sites by the presence or absence of another
factor). In these instances, motif analysis acts as a powerful tool to examine whether
differences in motif content is associated with membership of distinct
categories (e.g. motifs enriched at sites where both factors are bound vs singly
bound). Although the MEME Suite allows differential enrichment testing, it does
not inherently provide a mechanism for analyzing groups of sequences in
parallel, or performing motif analysis with any understanding of data
categories. The `memes` framework enables "data-aware" motif analysis workflows
by allowing named lists of `Biostrings` objects as input to each function. If
using `GenomicRanges`, users can split peak data on a metadata column using the
base R `split` function, and then use this result as input to `get_sequence`, which
will produce a list of `BioStrings` objects where each entry is named after the
data categories from the split column. When a list is used as input to a `memes`
function, it runs the corresponding MEME Suite tool for each object in the list.
Users can also pass the name(s) of a category to the `control` argument to
enable differential analysis of the remaining list members against the
control category sequences. Therefore, `memes` enables complex data-aware
differential motif analysis workflows using simple syntax to extend the
capabilities of the MEME Suite.

>>>>

Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers that can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, for example,
identification of kinase substrates based on similarities in phosphorylation
site sequence, or characterizing DNA elements based on shared transcription
factor binding sequences (TODO: ref kinase motif:
https://doi.org/10.1016/0968-0004(90)90073-K). Thus the ability to identify,
classify, and compare motifs represents a key step in the analysis of biological
sequences. **TODO: R/Bioconductor transition sentence**. Although tools that
implement these features exist in Bioconductor, they often rely on
proprietary data types or have poor interoperability with other tools, creating
barriers to comprehensive motif analysis within R. The recent addition of
`universalmotif` to the Bioconductor landscape partially addresss these issues
by standardizing motif formats and providing a powerful set of tools for
examining the properties of motifs. However the Bioconductor motif analysis
landscape remains incomplete \cite{universalmotif}.

The MEME Suite is a series of commandline and web-based tools that provide a variety
of features for motif analysis such as *de-novo* motif discovery, enrichment
testing, comparison, and matching \cite{meme-suite}. The MEME Suite is a widely
utilized set of tools, with over 22,000 citations to date, and greater than 30,000 unique
users of the webserver implementation annually \cite{meme-usage-report-page,
google-scholar-meme}. The MEME Suite family of tools provides a comprehensive
set of utilities to interrogate motif content across a broad range of biological
contexts. However, several factors limit the full potential of these tools for
use in data analysis. MEME Suite tools require carefully formatted inputs to
achieve full functionality, yet few tools exist to simplify the process of data
formatting, requiring instead that users write custom code to prepare their
data, or prepare the inputs by hand, both of which have the potential to be
error prone without rigorous testing. Furthermore, the output data from each
MEME Suite tool often have complex structures that must be parsed to extract the
full suite of information, again requiring users to write custom code for this
task. Finally, although the data-generation capabilities of the MEME Suite are
excellent, the tools lack powerful ways to visualize the results. Collectively,
these factors act as barriers to adoption, preclude deeper analysis of the
data, and limit communication of results to the scientific community.

@@

`memes` also provides simple utilities for data visualization where appropriate.
Providing R representations of MEME Suite data objects enables generation of
novel data visualizations using the well-established tools available in the R
landscape. Therefore, `memes` empowers users to deeply analyze and visualize
motif analysis data.

@@

MEME tools return HTML reports for each tool that display data in a
user-friendly way; however, these files are not ideal for downstream processing.
Depending on the tool, data are also returned in tab-separated, or XML format,
which are more amenable to computational processing. However, in the case that
tab-separated data are returned, results are often incomplete. For example, the
TomTom tool, which compares query motifs to a database of known motifs, returns
tab-separated results that do not contain the frequency matrix representations
of the matched motifs. Instead, users must write custom code to parse these
matrices back out from the input databases, creating additional barriers to
analysis. In the case that data are returned in XML format, these files contain
all relevant data; however, XML files are difficult to parse, again requiring users to
write custom code to extract the necessary data. Finally, the data types
contained in MEME Suite outputs are complex and multidimensional, and thus
require special data structures to properly organize the data in memory.

@@

Data returned from MEME Suite tools are complex and multidimensional, making
them difficult to represent in a simple data structure. For example, MEME and
DREME return de-novo discovered motifs from query sequences along with
statistical information about their enrichment. In this instance, a position
frequency matrix (PFM) of the discovered motif can be represented as a matrix,
while the properties of that matrix (e.g. the name of the motif, the E-value
from the enrichment test, background nucleotide frequency, etc.) must be encoded
outside of the matrix, while maintaining their relationship to the
corresponding PFM. However, other MEME Suite tools, such as TOMTOM, which
compares one motif to a series of several motifs to identify possible matches,
produce an additional layer of complexity such that input matrices (which
contain metadata as previously described) can have a one-to-many relationship
with other motif matrices (again with their own metadata). Thus an ideal
representation for these data is one that can hold an unlimited number of motif
matrices and their metadata, contain complex hierarchies, and be easily
manipulated using standard analysis tools.

The `universalmotif` data structure is a powerful R/Bioconductor representation
for motif matrices and their associated metadata \cite{universalmotif}. However,
`universalmotif` objects are not ideal for performing bulk manipulation or
queries of metadata, nor do they support complex data structures such as nested
elements. R data.frames support nesting arbitrary data types within columns, but
are not sufficient to properly store motif matrices. In order to create an
extensible, nested data structure for tracking MEME Suite data, `memes`
leverages the power of R data.frames and the `universalmotif` format by nesting
`universalmotif` objects within data.frames. These universalmotif data.frames
store information about motifs along rows, where each column stores metadata
associated with a motif, including a `motif` column holding a `universalmotif`
representation of that motif, which can be manipulated by altering columns of
the data.frame (described in detail below). Adding one-to-many relationships within this
structure is trivial, as additional unviersalmotif data.frames can be nested
within eachother. These universalmotif data.frames form the basis for a majority
of `memes` data outputs. These structures are also valid input types to `memes`
functions, and when used as such, output data are appended as new columns to the
input data, ensuring data provenance. Finally, because the universalmotif
data.frame is a standard R data.frame, it can be manipulated using base R or
`tidyverse` functions, lowering the cognitive overhead to interfacing with these
structures.

@@

Discriminative (or "differential") motif analysis, in which motifs are discovered in one set of
sequences relative to another set, can uncover biologically relevant motifs
associated with membership to distinct categories. For example, during analysis
of multiomic data, users can identify different functional categories of genomic
regions through integration with orthogonal datasets (e.g.  categorizing
transcription factor binding sites by the presence or absence of another
factor). In these instances, motif analysis acts as a powerful tool to examine whether
differences in motif content is associated with membership of distinct
categories (e.g. motifs enriched at sites where both factors are bound vs singly
bound). Although the MEME Suite allows differential enrichment testing, it does
not inherently provide a mechanism for analyzing groups of sequences in
parallel, or performing motif analysis with any understanding of data
categories. The `memes` framework enables "data-aware" motif analysis workflows
by allowing named lists of `Biostrings` objects as input to each function. If
using `GenomicRanges`, users can split peak data on a metadata column using the
base R `split` function, and then use this result as input to `get_sequence`, which
will produce a list of `BioStrings` objects where each entry is named after the
data categories from the split column. When a list is used as input to a `memes`
function, it runs the corresponding MEME Suite tool for each object in the list.
Users can also pass the name(s) of a category to the `control` argument to
enable differential analysis of the remaining list members against the
control category sequences. In this manner, `memes` enables complex data-aware
differential motif analysis workflows using simple syntax to extend the
capabilities of the MEME Suite.
